---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# gargle

<!-- badges: start -->
[![Build Status](https://travis-ci.org/r-lib/gargle.svg?branch=master)](https://travis-ci.org/r-lib/gargle)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/r-lib/gargle?branch=master&svg=true)](https://ci.appveyor.com/project/r-lib/gargle)
[![codecov.io](https://codecov.io/github/r-lib/gargle/coverage.svg?branch=master)](https://codecov.io/github/r-lib/gargle?branch=master)
[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

The goal of gargle is to take some of the agonizing pain out of working with Google APIs. This includes functions and classes for handling common credential types and for preparing, executing, and processing HTTP requests.

The target user of gargle is an *R package author* who is wrapping one of the ~250 Google APIs listed in the [APIs Explorer](https://developers.google.com/apis-explorer). gargle aims to play roughly the same role as [Google's official client libraries](https://developers.google.com/api-client-library/), but for R. gargle may also be useful to useRs making direct calls to Google APIs, who are prepared to navigate the details of low-level API access.

gargle's functionality falls into two main domains:

  * **Auth.** The `token_fetch()` function calls a series of concrete
    credential-fetching functions to obtain a valid access token (or it quietly
    dies trying).
    - This covers explicit service accounts, application default credentials,
      Google Compute Engine, and the standard OAuth2 browser flow.
    - gargle offers the `Gargle2.0` class, which extends `httr::Token2.0`. It is
      the default class for user OAuth 2.0 credentials. There are two main
      differences from `httr::Token2.0`: greater emphasis on the user's email
      (e.g. Google identity) and default token caching is at the user level.
  * **Requests and responses**. A family of functions helps to prepare HTTP
    requests, (possibly with reference to an API spec derived from a Discovery
    Document), make requests, and process the response.
    
See the [articles](https://gargle.r-lib.org/articles/) for holistic advice on how to use gargle.

## Installation

You can install the released version of gargle from [CRAN](https://CRAN.R-project.org) with:

```{r eval = FALSE}
install.packages("gargle")
```

And the development version from [GitHub](https://github.com/) with:

```{r eval = FALSE}
# install.packages("devtools")
devtools::install_github("r-lib/gargle")
```
## Basic usage

gargle is a low-level package and does not do anything visibly exciting on its own. But here's a bit of usage in an interactive scenario where a user confirms they want to use a specific Google identity and loads an OAuth2 token.

```{r eval = FALSE}
library(gargle)

token <- token_fetch()
#> The gargle package is requesting access to your Google account. Select a
#> pre-authorised account or enter '0' to obtain a new token. Press
#> Esc/Ctrl + C to abort.

#> 1: janedoe_personal@gmail.com
#> 2: janedoe@example.com

#> Selection: 1

token
#> <Token (via gargle)>
#>   <oauth_endpoint> google
#>              <app> gargle-demo
#>            <email> janedoe_personal@gmail.com
#>           <scopes> ...userinfo.email
#>      <credentials> access_token, expires_in, refresh_token, scope, ...
```

Here's an example of using request and response helpers to make a one-off request to the [Places API](https://developers.google.com/places/web-service/intro). We'll find the restaurants in a certain part of Vancouver, BC.

```{r}
library(gargle)

req <- request_build(
  method = "GET",
  path = "maps/api/place/nearbysearch/json",
  params = list(
    location = "49.268682,-123.167117",
    radius = 100,
    type = "restaurant"
  ),
  key = gargle_api_key(),
  base_url = "https://maps.googleapis.com"
)
resp <- request_make(req)
out <- response_process(resp)
vapply(out$results, function(x) x$name, character(1))
```

Please note that the 'gargle' project is released with a [Contributor Code of Conduct](.github/CODE_OF_CONDUCT.md). By contributing to this project, you agree to abide by its terms.
