---
title: "Discovery Document Ingest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Discovery Document Ingest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(gargle)
```

gargle provides support for creating and making HTTP requests via these functions:

  * `request_develop(endpoint, params, base_url)`: a.k.a. "The Smart One". This function processes the info in `params` relative to detailed knowledge about the `endpoint`, derived from an API Discovery Document.
  * `request_build(method, path, params, body, token, key, base_url)`: a.k.a. The Dumb One". This function does purely mechanical operations to finalize a request.
  * `request_make(x, ..., user_agent)`: Consults `x$method` to determine which `httr::VERB()` to call, then calls it with the rest of `x` and `...` passed as arguments.
  
They are usually called in the above order, though they don't have to be used that way.

It is also fine to ignore this part of gargle and use it only for help with auth. They are separate parts of the package.

## What the payoff?

Why would the developer of a Google-API-wrapping package care about the request helpers in gargle?

You can get some safety and concision in your code, for a modest investment.

Hundreds of Google APIs -- the ones addressed by the [API Discovery Service](https://developers.google.com/discovery/) -- have a lot of shared behaviour. By ingesting the metadata provided by this service, you can maintain less code in your package, while also decreasing the chance that you and your users submit ill-formed requests.

The request helpers in gargle check the combined inputs from user and developer against API metadata:

  * If required parameters are missing, an error is thrown.
  * If unrecognized parameters are submitted, an error is thrown.
  * Parameters are automatically placed in their correct location: URL substitution, query, or body.
  * *Is there something else you care about? It is possible to do more, but it would help to have concrete requests.*
  
Google provides [API libraries for several languages](https://developers.google.com/api-client-library/), including Java, Go, Python, JavaScript, Ruby and more (but not R). However, note that all of these machine-generated libraries are based on the API Discovery Service. It is the [official recommendation](https://developers.google.com/discovery/v1/using#build) to use the Discovery Document when building client libraries. The gargle package aims to implement this in a way that is idiomatic for R and its developers.
  
## Discovery Documents

Google's [API Discovery Service](https://developers.google.com/discovery/) "provides a lightweight, JSON-based API that exposes machine-readable metadata about Google APIs". We recommend ingesting this metadata into an R list, stored as internal data in an API-wrapping client package. Then, HTTP requests inside high-level functions can be made concisely and safely, by referring to this metadata. The combined use of this data structure and gargle's request helpers can eliminate a lot of boilerplate data and logic that are shared across Google APIs and across endpoints within an API.

The gargle package ships with some functions and scripts to facilitate the ingest of a Discovery Document. These files are located here:

```{r}
ddi_dir <- system.file("discovery-doc-ingest", package = "gargle")
list.files(ddi_dir)
```

Main files of interest to the developer of a client package:

  * `ingest-functions.R` is a collection of functions for downloading and ingesting a Discovery Document.
  * `drive-example.R` uses those functions to ingest metadata on the Drive v3 API and store it as an internal data object for use in [googledrive](https://googledrive.tidyverse.org).
  
The remaining files present some analysis of the Discovery Document for the Discovery API itself (very meta!) and write some files we consult for reference.
  
Why aren't the ingest functions exported by gargle? First, we regard this as functionality that is needed at development time, not install or run time. This is something you'll do every few months, probably associated with preparing a release of a wrapper package. Second, the packages that are useful for wrangling JSON and lists are not existing dependencies of gargle, so putting these function in gargle would require some unappealing compromises.

## Method (or endpoint) data

Our Discovery Document ingest process leaves you with an R list. Let's assume it's available in your package's namespace as an internal object named `.endpoints`. Each item represents one method of the API (Google's vocabulary) or an endpoint (term I originally used in, e.g., googledrive ... maybe I should conform to Google?).

Each endpoint has an `id`. These `id`s are also used as names for the list. Examples of some `id`s from the Drive and Sheets APIs:

```
drive.about.get
drive.files.create
drive.teamdrives.list
sheets.spreadsheets.create
sheets.spreadsheets.values.clear
sheets.spreadsheets.sheets.copyTo
```

Retrieve the metadata for one endpoint by name, e.g.:

```{r, eval = FALSE}
.endpoints[["drive.files.create"]]
```

That info can be passed along to `request_develop(endpoint, params, base_url)`, which conducts sanity checks and combines this external knowledge with the data coming from the user and developer via `params`.

## Helpers to create requests

Here's the model we used in googledrive. There is a low-level request helper, `googledrive::request_generate()`, that is used to form every request in the package. It is exported as part of a low-level API for expert use, but most users will never know it exists.

```{r eval = FALSE}
# googledrive::
request_generate <- function(endpoint = character(),
                             params = list(),
                             key = NULL,
                             token = drive_token()) {
  ept <- .endpoints[[endpoint]]
  if (is.null(ept)) {
    stop_glue("\nEndpoint not recognized:\n  * {endpoint}")
  }

  ## modifications specific to googledrive package
  params$key <- key %||% params$key %||% drive_api_key()
  if (!is.null(ept$parameters$supportsTeamDrives)) {
    params$supportsTeamDrives <- TRUE
  }

  req <- gargle::request_develop(endpoint = ept, params = params)
  gargle::request_build(
    path = req$path,
    method = req$method,
    params = req$params,
    body = req$body,
    token = token
  )
}
```

The `endpoint` argument specifies an endpoint by its name, a.k.a. its `id`.

`params` is where the processed user input goes.

`key` and `token` refer to an API key and OAuth2 token, respectively. Both can be populated by default, but it is possible to pass them explicitly. This is also where we enforce the rule that we always send exactly **one** of `key` and `token`.

`googledrive::request_generate()` takes care of everything specific to the Drive API and the user's input and task. Then it uses `gargle::request_develop()` followed by `gargle::request_build()` to finish preparing the request.

## Helper to make requests

The output of `gargle::request_build()` specifies an HTTP request.

`gargle::request_make()` can be used to actually execute it.

```{r, eval = FALSE}
# gargle::
request_make <- function(x, ..., user_agent = gargle_user_agent()) {
  stopifnot(is.character(x$method))
  method <- switch(
    x$method,
    GET = httr::GET,
    POST = httr::POST,
    PATCH = httr::PATCH,
    PUT = httr::PUT,
    DELETE = httr::DELETE,
    stop_glue("Not a recognized HTTP method: {bt(m)}", m = x$method)
  )
  method(
    url = x$url,
    body = x$body,
    x$token,
    user_agent,
    ...
  )
}
```

`request_make()` consults `x$method` to identify the `httr::VERB()` and then calls it with the remainder of `x`, `...` and the `user_agent`.

In googledrive we have a thin wrapper around this that injects the googledrive user agent:

```{r, eval = FALSE}
# googledrive::
request_make <- function(x, ...) {
  gargle::request_make(x, ..., user_agent = drive_ua())
}
```
