---
title: "Troubleshooting gargle auth"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Troubleshooting gargle auth}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(gargle)
```

## `gargle_quiet`

There is a package-wide option that controls gargle's verbosity: `gargle_quiet`. The function `gargle_quiet()` reveals the current value:

```{r}
gargle_quiet()
```

It defaults to `TRUE`, i.e. gargle defaults to being very quiet. This is because gargle is designed to try a bunch of auth methods (many of which will fail) and persist doggedly until one succeeds. If none succeeds, gargle tries to guide the user through auth or, in a non-interactive session, it throws an error.

If you need to see all those gory details, toggle the `gargle_quiet` option to `FALSE` and you'll get much more output as gargle works through various auth approaches.

```{r}
# save current value
op <- options(gargle_quiet = FALSE)

gargle_quiet()

# restore original value
options(op)
```

## `gargle_oauth_sitrep()`

`gargle_oauth_sitrep()` provides an OAuth2 "situation report".

`gargle_oauth_sitrep()` is only relevant to OAuth2 user tokens. If you are using (or struggling to use) a service account token, Application Default Credentials, or credentials from the GCE metadata service, `gargle_oauth_sitrep()` isn't going to help you figure out what's going on.

Here is indicative output of `gargle_oauth_sitrep()`, for someone who has accepted the default OAuth cache location and has played with several APIs via gargle-using packages.

```{r, eval = FALSE}
gargle_oauth_sitrep()
#' gargle OAuth cache path:
#' /Users/janedoe/.R/gargle/gargle-oauth
#' 
#' 14 tokens found
#' 
#' email                         app         scope                          hash...   
#' ----------------------------- ----------- ------------------------------ ----------
#' abcdefghijklm@gmail.com       thingy      ...bigquery, ...cloud-platform 128f9cc...
#' buzzy@example.org             gargle-demo                                15acf95...
#' stella@example.org            gargle-demo ...drive                       4281945...
#' abcdefghijklm@gmail.com       gargle-demo ...drive                       48e7e76...
#' abcdefghijklm@gmail.com       tidyverse                                  69a7353...
#' nopqr@ABCDEFG.com             tidyverse   ...spreadsheets.readonly       86a70b9...
#' abcdefghijklm@gmail.com       tidyverse   ...drive                       d9443db...
#' nopqr@HIJKLMN.com             tidyverse   ...drive                       d9443db...
#' nopqr@ABCDEFG.com             tidyverse   ...drive                       d9443db...
#' stuvwzyzabcd@gmail.com        tidyverse   ...drive                       d9443db...
#' efghijklmnopqrtsuvw@gmail.com tidyverse   ...drive                       d9443db...
#' abcdefghijklm@gmail.com       tidyverse   ...drive.readonly              ecd11fa...
#' abcdefghijklm@gmail.com       tidyverse   ...bigquery, ...cloud-platform ece63f4...
#' nopqr@ABCDEFG.com             tidyverse   ...spreadsheets                f178dd8...
```

It is relatively harmless to delete the folder serving as the OAuth cache. Or, if you have reason to believe one specific cached token is causing you pain, you could delete a specific token (an `.rds` file) from the cache. OAuth user tokens are meant to be perishable and replaceable.

If you choose to delete your cache (or a specific token), here is the fallout you can expect:

  * You will need to re-auth (usually, meaning the browser dance) in projects
    that have been using the deleted tokens.
  * If you have `.R` or `.Rmd` files that you execute or render
    non-interactively, presumably with code such as
    `PKG_auth(email = "janedoe@example.com")`, those won't run non-interactively
    until you've obtained and cached a token for the package and that
    identity (email) interactively once.
  * A specific Google user (email) can only have a certain number of OAuth
    tokens at a time (something like 50 per OAuth client). So, whenever you get
    a new token (as opposed to refreshing an existing token), there is the
    potential for it to invalidate an older token. This is unlikely to have any
    practical effect for a normal user, but can be noticeable for someone
    developing against a Google API or someone working from many different
    machines / caches.
 *  If you have rigged some remote mission critical thing (e.g. a Shiny app or
    cron job) to use a user OAuth token (which is not a great idea), one day
    continued acquisition of new tokens in your normal interactive life will
    invalidate the mission critical token. Your thing (the Shiny app or cron
    job) will mysteriously fail because the OAuth token can't be refreshed. Be
    prepared to deal with that periodically or, better yet, upgrade to a more
    robust strategy for [non-interactive auth](https://gargle.r-lib.org/articles/non-interactive-auth.html).
